### 2025-07-12 factor-analysis.qmd
20250712 - add interpretation

@@

Factor analysis involves observed (manifest) variables and unobserved (latent) factors.\index{factor analysis}\index{latent variable}
Factor analysis assumes that the latent factor influences the manifest variables, and the latent factor therefore reflects the common variance among the variables.\index{factor analysis}\index{latent variable}\index{latent variable}
A factor model potentially includes factor loadings, residuals (errors or disturbances), intercepts/means, covariances, and regression paths.\index{factor analysis}
When depicting a factor analysis model, rectangles represent variables we observe (i.e., manifest variables), and circles represent latent (i.e., unobserved) variables.\index{factor analysis}\index{latent variable}
A regression path indicates a hypothesis that one variable (or factor) influences another, and it is depicted using a single-headed arrow.\index{factor analysis}
The standardized regression coefficient (i.e., beta or $\beta$) represents the strength of association between the variables or factors.\index{factor analysis}\index{standardized regression coefficient}
A factor loading is a regression path from a latent factor to an observed (manifest) variable.\index{factor analysis}\index{factor analysis!factor loading}\index{latent variable}
The standardized factor loading represents the strength of association between the variable and the latent factor.\index{factor analysis}\index{factor analysis!factor loading}\index{latent variable}\index{standardized regression coefficient}
A residual is variance in a variable (or factor) that is unexplained by other variables or factors.\index{factor analysis}\index{factor analysis!residual}
A variable's intercept is the expected value of the variable when the factor(s) (onto which it loads) is equal to zero.\index{factor analysis}\index{factor analysis!intercept}
A covariance is the unstandardized index of the strength of association between between variables (or factors), and it is depicted with a double-headed arrow.\index{factor analysis}\index{factor analysis!covariance}
Because a covariance is unstandardized, its scale depends on the scale of the variables.\index{factor analysis!covariance}
A covariance path between two variables represents omitted shared cause(s) of the variables.\index{factor analysis}\index{factor analysis!covariance}
For instance, if you depict a covariance path between two variables, it means that there is a shared cause of the two variables that is omitted from the model (for instance, if the common cause is not known or was not assessed).\index{factor analysis}\index{factor analysis!covariance}

@@

When the latent factors are uncorrelated, the (standardized) error term for an indicator is calculated as 1 minus the sum of squared standardized factor loadings for a given item (including cross-loadings).\index{factor analysis}\index{factor analysis!residual}\index{factor analysis!factor loading}\index{cross-loading}
A cross-loadings is when a variable loads onto more than one latent factor.\index{factor analysis}\index{cross-loading}

@@

## Exploratory Factor Analysis {#sec-efa}

@@

### Factor Rotation {#sec-factorAnalysisRotation}

@@

### Example {#sec-efaExample}

@@

We generated the scree plot in @fig-efaScreePlot1 using the `fa.parallel()` function of the `psych` package [@R-psych].

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-efaVSSPlot1] and [-@fig-efaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].

@@

@@

A path diagram of the EFA model in @fig-efaPathDiagram-3factor was created using the `lavaanPlot()` function of the `lavaanPlot` package [@R-lavaanPlot].

@@

Here is the syntax for estimating a three-factor EFA using exploratory structural equation modeling (ESEM).
Estimating the model in a ESEM framework allows us to make modifications to the model, such as adding correlated residuals, and adding predictors or outcomes of the latent factors.

@@

@@

@@

@@

Here is a modification of the 3-factor model with correlated residuals:

@@

@@

@@

@@

## Confirmatory Factor Analysis {#sec-cfa}

### Example {#sec-cfaExample}

@@

We fit CFA models using the `cfa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].

@@

@@

@@

@@

@@

@@

```{r}
lavaan::fitMeasures(
  cfa3factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

@@

@@

@@

@@

@@

## Determining the Number of Factors to Retain {#sec-factorAnalysisNumFactors}

A goal of factor analysis and [principal component analysis](#sec-pca) is simplification or parsimony, while still explaining as much variance as possible.\index{parsimony}\index{factor analysis}\index{principal component analysis}
The hope is that you can have fewer factors that explain the associations between the variables than the number of observed variables.\index{factor analysis}
But how do you decide on the number of factors?\index{factor analysis}

There are more sophisticated ways of using a scree plot, but they usually end up at a similar decision.\index{factor analysis}\index{scree plot}
Examples of more sophisticated tests include parallel analysis and very simple structure (VSS) plots.\index{factor analysis}\index{very simple structure plot}\index{parallel analysis}
In a parallel analysis, you examine where the eigenvalues from observed data and random data converge, so you do not retain a factor that explains less variance than would be expected by random chance.\index{factor analysis}\index{parallel analysis}\index{eigenvalue}

In general, my recommendation is to use Cattell's scree test, and then test the factor solutions with plus or minus one factor.\index{factor analysis}\index{scree plot}
You should never accept factors with eigenvalues less than zero (or components from [principal component analysis](#sec-pca) with eigenvalues less than one), because they are likely to be largely composed of error.\index{factor analysis}\index{eigenvalue}
If you are using maximum likelihood factor analysis, you can compare the fit of various models with model fit criteria to see which model fits best for its parsimony.\index{parsimony}\index{factor analysis}
A model will always fit better when you add additional parameters or factors, so you examine if there is *significant* improvement in model fit when adding the additional factor—that is, we keep adding complexity until additional complexity does not buy us much.\index{parsimony}\index{factor analysis}
Always try a factor solution that is one less and one more than suggested by Cattell's scree test to buffer your final solution because the purpose of factor analysis is to explain things and to have interpretability.\index{factor analysis}\index{scree plot}
Even if all rules or indicators suggest to keep X number of factors, maybe $\pm$ one factor helps clarify things.\index{factor analysis}
Even though factor analysis is empirical, theory and interpretatability should also inform decisions.\index{factor analysis}

## Interpreting and Using Latent Factors {#sec-interpretingAndUsingLatentFactors}

The next step is interpreting the model and latent factors.\index{factor analysis}
One data matrix can lead to many different (correct) models—you must choose one based on the factor structure and theory.\index{factor analysis}
Use theory to interpret the model and label the factors.\index{factor analysis}

>>>>

> "All models are wrong, but some are useful."
>
> --- George Box [-@Box1979, p. 202]

@@

Factor analysis involves observed (manifest) variables and unobserved (latent) factors.\index{factor analysis}\index{latent variable}
Factor analysis assumes that the latent factor influences the manifest variables, and the latent factor therefore reflects the common variance among the variables.\index{factor analysis}\index{latent variable}\index{latent variable}
A factor model potentially includes factor loadings, residuals (errors or disturbances), intercepts/means, covariances, and regression paths.\index{factor analysis}
When depicting a factor analysis model, rectangles represent variables we observe (i.e., manifest variables), and circles represent latent (i.e., unobserved) variables.\index{factor analysis}\index{latent variable}
A regression path indicates a hypothesis that one variable (or factor) influences another, and it is depicted using a single-headed arrow.\index{factor analysis}
The standardized regression coefficient (i.e., beta or $\beta$) represents the strength of association between the variables or factors.\index{factor analysis}\index{standardized regression coefficient}
A factor loading is a regression path from a latent factor to an observed (manifest) variable.\index{factor analysis}\index{factor analysis!factor loading}\index{latent variable}
The standardized factor loading represents the strength of association between the variable and the latent factor, where conceptually, it is intended to reflect the magnitude that the latent factor influences the observed variable.\index{factor analysis}\index{factor analysis!factor loading}\index{latent variable}\index{standardized regression coefficient}
A residual is variance in a variable (or factor) that is unexplained by other variables or factors.\index{factor analysis}\index{factor analysis!residual}
A variable's intercept is the expected value of the variable when the factor(s) (onto which it loads) is equal to zero.\index{factor analysis}\index{factor analysis!intercept}
A [*covariance*](#sec-correlationExamplesCovariance) is the unstandardized index of the strength of association between between two variables (or factors), and it is depicted with a double-headed arrow.\index{factor analysis}\index{factor analysis!covariance}

Because a covariance is unstandardized, its scale depends on the scale of the variables.
The covariance between two variables is the average product of their deviations from their respective means, as in @eq-covariance.
The covariance of a variable with itself is equivalent to its variance, as in @eq-covarianceToVariance.
By contrast, a [*correlation*](#sec-correlation) is a standardized index of the strength of association between two variables.
Because a correlation is standardized (fixed between [−1,1]), its scale does not depend on the scales of the variables.
Because a covariance is unstandardized, its scale depends on the scale of the variables.\index{factor analysis!covariance}
A covariance path between two variables represents omitted shared cause(s) of the variables.\index{factor analysis}\index{factor analysis!covariance}
For instance, if you depict a covariance path between two variables, it means that there is a shared cause of the two variables that is omitted from the model (for instance, if the common cause is not known or was not assessed).\index{factor analysis}\index{factor analysis!covariance}

@@

When the latent factors are uncorrelated, the (standardized) error term for an indicator is calculated as 1 minus the sum of squared standardized factor loadings for a given item (including cross-loadings).\index{factor analysis}\index{factor analysis!residual}\index{factor analysis!factor loading}\index{cross-loading}
A cross-loading is when a variable loads onto more than one latent factor.\index{factor analysis}\index{cross-loading}

@@

## Path Diagrams {#sec-pathDiagrams}

A key tool when designing a structural equation model is a conceptual depiction of the hypothesized causal processes.
A path diagram depicts the hypothesized causal processes that link two or more variables.
@Karch2025_packages provides a tool to create `lavaan` `R` syntax from a path analytic diagram: <https://lavaangui.org>.

In a path analysis diagram, rectangles represent variables we observe, and circles represent latent (i.e., unobserved) variables.
Single-headed arrows indicate regression paths, where conceptually, one variable is thought to influence another variable.
Double-headed arrows indicate covariance paths, where conceptually, two variables are associated for some unknown reason (i.e., an omitted shared cause).

## Decisions in Factor Analysis {#sec-factorAnalysisDecisions}

There are five primary decisions to make in factor analysis:

1. what variables to include in the model and how to scale them
1. method of factor extraction: whether to use exploratory or confirmatory factor analysis
1. if using exploratory factor analysis, whether and how to rotate factors
1. how many factors to retain (and what variables load onto which factors)
1. how to interpret and use the factors

The answer you get can differ highly depending on the decisions you make.
Below, we provide guidance for each of these decisions.

### 1. Variables to Include and their Scaling {#sec-variablesToInclude-factorAnalysis}

The first decision when conducting a factor analysis is which variables to include and the scaling of those variables.
What factors you extract can differ widely depending on what variables you include in the analysis.
For example, if you include many variables from the same source (e.g., self-report), it is possible that you will extract a factor that represents the common variance among the variables from that source (i.e., the self-reported variables).
This would be considered a method factor, which works against the goal of estimating latent factors that represent the constructs of interest (as opposed to the measurement methods used to estimate those constructs).

An additional consideration is the scaling of the variables: whether to use raw variables, standardized variables, or dividing some variables by a constant to make the variables' variances more similar.
Before performing a [principal component analysis](#sec-pca) (PCA), it is generally important to ensure that the variables included in the PCA are on the same scale.
PCA seeks to identify components that explain variance in the data, so if the variables are not on the same scale, some variables may contribute considerably more variance than others.
A common way of ensuring that variables are on the same scale is to standardize them using, for example,  [*z*-scores](#zScores) that have a mean of zero and standard deviation of one.
By contrast, factor analysis can better accommodate variables that are on different scales.

### 2. Method of Factor Extraction {#sec-methodOfFactorExtraction}

#### Exploratory Factor Analysis {#sec-efa}

@@

#### Confirmatory Factor Analysis {#sec-cfa}

### 3. Factor Rotation {#sec-factorAnalysisRotation}

@@

### 4. Determining the Number of Factors to Retain {#sec-factorAnalysisNumFactors}

A goal of factor analysis and [principal component analysis](#sec-pca) is simplification or parsimony, while still explaining as much variance as possible.\index{parsimony}\index{factor analysis}\index{principal component analysis}
The hope is that you can have fewer factors that explain the associations between the variables than the number of observed variables.\index{factor analysis}
It does not make sense to replace 18 variables with 18 latent factors because that would not result in any simplification.
But how do you decide on the number of factors?\index{factor analysis}

There are more sophisticated ways of using a scree plot, but they usually end up at a similar decision.\index{factor analysis}\index{scree plot}
Examples of more sophisticated tests include parallel analysis and very simple structure (VSS) plots.\index{factor analysis}\index{very simple structure plot}\index{parallel analysis}
In a parallel analysis, you examine where the eigenvalues from observed data and random data converge, so you do not retain a factor that explains less variance than would be expected by random chance.\index{factor analysis}\index{parallel analysis}\index{eigenvalue}
Using the VSS criterion, the optimal number of factors to retain is the number of factors that maximizes the VSS criterion [@Revelle1979].
The VSS criterion is evaluated with models in which factor loadings for a given item that are less than the maximum factor loading for that item are suppressed to zero, thus forcing simple structure (i.e., no cross-loadings).
The goal is finding a factor structure with interpretability so that factors are clearly distinguishable.
Thus, we want to identify the number of factors with the highest VSS criterion.

In general, my recommendation is to use Cattell's scree test, and then test the factor solutions with plus or minus one factor, in addition to examining [model fit](#sec-modelFitIndices).\index{factor analysis}\index{scree plot}
You should never accept factors with eigenvalues less than zero (or components from [PCA](#sec-pca) with eigenvalues less than one), because they are likely to be largely composed of error.\index{factor analysis}\index{eigenvalue}
If you are using maximum likelihood factor analysis, you can compare the fit of various models with model fit criteria to see which model fits best for its parsimony.\index{parsimony}\index{factor analysis}
A model will always fit better when you add additional parameters or factors, so you examine if there is *significant* improvement in model fit when adding the additional factor—that is, we keep adding complexity until additional complexity does not buy us much.\index{parsimony}\index{factor analysis}
Always try a factor solution that is one less and one more than suggested by Cattell's scree test to buffer your final solution because the purpose of factor analysis is to explain things and to have interpretability.\index{factor analysis}\index{scree plot}
Even if all rules or indicators suggest to keep X number of factors, maybe $\pm$ one factor helps clarify things.\index{factor analysis}
Even though factor analysis is empirical, theory and interpretatability should also inform decisions.\index{factor analysis}

#### Model Fit Indices {#sec-modelFitIndices}

In factor analysis, we fit a model to observed data, or to the variance-covariance matrix, and we evaluate the degree of model misfit.
That is, fit indices evaluate how likely it is that a given causal model gave rise to the observed data.
Various model fit indices can be used for evaluating how well a model fits the data and for comparing the fit of two competing models.
Fit indices known as absolute fit indices compare whether the model fits better than the best-possible fitting model (i.e., a saturated model).
Examples of absolute fit indices include the chi-square test, root mean square error of approximation (RMSEA), and the standardized root mean square residual (SRMR).

The chi-square test evaluates whether the model has a significant degree of misfit relative to the best-possible fitting model (a saturated model that fits as many parameters as possible; i.e., as many parameters as there are degrees of freedom); the null hypothesis of a chi-square test is that there is no difference between the predicted data (i.e., the data that would be observed if the model were true) and the observed data.
Thus, a non-significant chi-square test indicates good model fit.
However, because the null hypothesis of the chi-square test is that the model-implied covariance matrix is exactly equal to the observed covariance matrix (i.e., a model of perfect fit), this may be an unrealistic comparison.
Models are simplifications of reality, and our models are virtually never expected to be a perfect description of reality.
Thus, we would say a model is "useful" and partially validated if "it helps us to understand the relation between variables and does a 'reasonable' job of matching the data...A perfect fit may be an inappropriate standard, and a high chi-square estimate may indicate what we already know—that the hypothesized model holds approximately, not perfectly." [@Bollen1989, p. 268].
The power of the chi-square test depends on sample size, and a large sample will likely detect small differences as significantly worse than the best-possible fitting model [@Bollen1989].

RMSEA is an index of absolute fit.
Lower values indicate better fit.

SRMR is an index of absolute fit with no penalty for model complexity.
Lower values indicate better fit.

There are also various fit indices known as incremental, comparative, or relative fit indices that compare whether the model fits better than the worst-possible fitting model (i.e., a "baseline" or "null" model).
Incremental fit indices include a chi-square difference test, the comparative fit index (CFI), and the Tucker-Lewis index (TLI).
Unlike the chi-square test comparing the model to the best-possible fitting model, a significant chi-square test of the relative fit index indicates better fit—i.e., that the model fits better than the worst-possible fitting model.

CFI is another relative fit index that compares the model to the worst-possible fitting model.
Higher values indicate better fit.

TLI is another relative fit index.
Higher values indicate better fit.

Parsimony fit include fit indices that use information criteria fit indices, including the Akaike Information Criterion (AIC) and the Bayesian Information Criterion (BIC).\index{parsimony}
BIC penalizes model complexity more so than AIC.
Lower AIC and BIC values indicate better fit.

Chi-square difference tests and CFI can be used to compare two nested models.
AIC and BIC can be used to compare two non-nested models.

Criteria for acceptable fit and good fit of SEM models are in @semFitIndices.
In addition, dynamic fit indexes have been proposed based on simulation to identify fit index cutoffs that are tailored to the characteristics of the specific model and data [@McNeish2023]; dynamic fit indexes are available via the `dynamic` package [@R-dynamic] or with a [webapp](https://dynamicfit.app).

| SEM Fit Index | Acceptable Fit | Good Fit   |
|---------------|----------------|------------|
| RMSEA         | $\leq$ .08     | $\leq$ .05 |
| CFI           | $\geq$ .90     | $\geq$ .95 |
| TLI           | $\geq$ .90     | $\geq$ .95 |
| SRMR          | $\leq$ .10     | $\leq$ .08 |

: Criteria for Acceptable and Good Fit of Structural Equation Models Based on Fit Indices. {#tbl-semFitIndices}

However, good model fit does not necessarily indicate a true model.

In addition to global fit indices, it can also be helpful to examine evidence of local fit, such as the residual covariance matrix.
The residual covariance matrix represents the difference between the observed covariance matrix and the model-implied covariance matrix (the observed covariance matrix minus the model-implied covariance matrix).
These difference values are called *covariance residuals*.
Standardizing the covariance matrix by converting each to a correlation matrix can be helpful for interpreting the magnitude of any local misfit.
This is known as a residual correlation matrix, which is composed of *correlation residuals*.
Correlation residuals greater than |.10| are possible evidence for poor local fit [@Kline2023].
If a correlation residual is positive, it suggests that the model underpredicts the observed association between the two variables (i.e., the observed covariance is greater than the model-implied covariance).
If a correlation residual is negative, it suggests that the model overpredicts their observed association between the two variables (i.e., the observed covariance is smaller than the model-implied covariance).
If the two variables are connected by only indirect pathways, it may be helpful to respecify the model with direct pathways between the two variables, such as a direct effect (i.e., regression path) or a covariance path.
For guidance on evaluating local fit, see @Kline2024.

### 5. Interpreting and Using Latent Factors {#sec-interpretingAndUsingLatentFactors}

The next step is interpreting the model and latent factors.\index{factor analysis}
One data matrix can lead to many different (correct) models—you must choose one based on the factor structure and theory.\index{factor analysis}
Use theory to interpret the model and label the factors.\index{factor analysis}
In latent variable models, factors have meaning.
You can use them as predictors, mediators, moderators, or outcomes.
When possible, it is preferable to use the factors by examining their associations with other variables in the same model.
You can extract factor scores, if necessary, for use in other analyses; however, it is preferable to examine the associations between factors and other variables in the same model if possible.
And, using latent factors helps disattenuate associations for measurement error, to identify what the association is between variables when removing random measurement error.

## Example of Exploratory Factor Analysis {#sec-efaExample}

@@

We generated the scree plot in @fig-efaScreePlot1 using the `fa.parallel()` function of the `psych` package [@R-psych].
The optimal coordinates and the acceleration factor attempt to operationalize the Cattell scree test: i.e., the "elbow" of the scree plot [@Ruscio2012].
The optimal coordinators factor is quantified using a series of linear equations to determine whether observed eigenvalues exceed the predicted values.
The acceleration factor is quantified using the acceleration of the curve, that is, the second derivative.
The Kaiser-Guttman rule states to keep principal components whose eigenvalues are greater than 1.
However, for [exploratory factor analysis](#efa) (as opposed to [PCA](#pca)), the criterion is to keep the factors whose eigenvalues are greater than zero (i.e., not the factors whose eigenvalues are greater than 1) [@Dinno2014].

The number of factors to keep would depend on which criteria one uses.
Based on the rule to keep factors whose eigenvalues are greater than zero and based on the parallel test, we would keep five factors.
However, based on the Cattell scree test (the "elbow" of the screen plot minus one), we would keep three factors.
If using the optimal coordinates, we would keep eight factors; if using the acceleration factor, we would keep one factor.
Therefore, interpretability of the factors would be important for deciding how many factors to keep.

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-efaVSSPlot1] and [-@fig-efaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].
In addition to VSS plots, the output also provides additional criteria by which to determine the optimal number of factors, each for which lower values are better, including the Velicer minimum average partial (MAP) test, the Bayesian information criterion (BIC), the sample size-adjusted BIC (SABIC), and the root mean square error of approximation (RMSEA).
Depending on the criterion, the optimal number of factors extracted varies between 2 and 8 factors.

@@

The model fits well according to CFI with 5 or more factors; the model fits well according to RMSEA with 6 or more factors.
However, 5+ factors does not represent much of a simplification (relative to the 18 variables included).
Moreover, in the model with four factors, only one variable had a significant loading on Factor 1.
Thus, even with four factors, one of the factors does not seem to represent the aggregation of multiple variables.
For these reasons, and because the "elbow test" for the scree plot suggested three factors, we decided to retain three factors and to see if we could achieve better fit by making additional model modifications (e.g., correlated residuals).
Correlated residuals may be necessary, for example, when variables are correlated for reasons other than the latent factors.

@@

A path diagram of the three-factor EFA model in @fig-efaPathDiagram-3factor was created using the `lavaanPlot()` function of the `lavaanPlot` package [@R-lavaanPlot].

@@

Here is the syntax for estimating a three-factor EFA using exploratory structural equation modeling (ESEM).
Estimating the model in a ESEM framework allows us to make modifications to the model, such as adding correlated residuals, and adding predictors or outcomes of the latent factors. 
The syntax below represents the same model (with the same fit indices) as the three-factor EFA model above.

@@

The fit indices suggests that the model does not fit well to the data and that additional model modifications are necessary.
The fit indices are below.

@@

We can examine the model modification indices to identify parameters that, if estimated, would substantially improve model fit.
For instance, the modification indices below indicate additional correlated residuals that could substantially improve model fit.
However, it is generally not recommended to blindly estimate additional parameters solely based on modification indices, which can lead to data dredging and [overfitting](#sec-overfitting).
Rather, it is generally advised to consider modification indices in light of theory.
Based on the modification indices, we will add several correlated residuals to the model, to help account for why variables are associated with each other for reasons other than their underlying latent factors.

@@

Below are factor scores from the model for the first six players:

@@

Below is a modification of the three-factor model with correlated residuals.
For instance, it makes sense that passing completions and attempts are related to each other (even after accounting for their latent factor).

@@

The model fits substantially better (though not perfectly) with the additional correlated residuals.
Below are the model fit indices:

@@

Model fit of nested models can be compared with a chi-square difference test.
This allows us to evaluate whether the more complex model fits signficantly better than the less complex model.
In this case, our more complex model is the three-factor model with correlated residuals; the less complex model is the three-factor model without correlated residuals.
The modified model with the correlated residuals and the original model are considered "nested" models.
The original model is nested within the modified model because the modified model includes all of the terms of the original model along with additional terms.

@@

In this case, the model with the correlated residuals fit significantly better (i.e., has a significantly smaller chi-square value) than the model without the correlated residuals.

@@

## Example of Confirmatory Factor Analysis {#sec-cfaExample}

@@

We fit CFA models using the `cfa()` function of the `lavaan` package [@Rosseel2012_packages; @R-lavaan].
We compare one-, two-, and three-factor CFA models.

Below is the syntax for the one-factor CFA model:

@@

The one-factor model did not fit well according to the fit indices:

@@

Below is the syntax for the two-factor CFA model:

@@

The two-factor model did not fit well according to the fit indices:

@@

Because the one-factor model is nested within the two-factor model, we can compare them using a chi-square difference test.
The two factor model fit considerably better than the one-factor model in terms of a lower chi-square value:

@@

Below is the syntax for the three-factor model:

@@

The three-factor model did not fit well according to fit indices:
```{r}
lavaan::fitMeasures(
  cfa3factor_fit,
  fit.measures = c(
    "chisq", "df", "pvalue",
    "chisq.scaled", "df.scaled", "pvalue.scaled",
    "chisq.scaling.factor",
    "baseline.chisq","baseline.df","baseline.pvalue",
    "rmsea", "cfi", "tli", "srmr",
    "rmsea.robust", "cfi.robust", "tli.robust"))
```

However, we know that the three-factor model fit improved considerably when accounting for correlated residuals.
So, we plan to examine modification indices to see if we can account for covariances between variables that were not explained by their underlying latent factors.

@@

Below are the modification indices, suggesting potential model modifications that would improve model fit such as correlated residuals and cross-loadings.

@@

Because the two-factor model is nested within the three-factor model, we can compare them using a chi-square difference test.
The three factor model fit considerably better than the two-factor model in terms of a lower chi-square value:

@@

Based on the model modifications suggested by the modification indices for the three-factor model, we modify the model to account for correlated residuals, using the syntax below:

@@

The three-factor model with correlated residuals fit was acceptable according to CFI and TLI.
The model fit was subpar for RMSEA and SRMR.

@@

### 2025-07-12 index.qmd
20250712 - add interpretation

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "corrplot", "mgcv", "rms", 
  "car", "lavaan", "lavaangui", "mice", "miceadds", "interactions", 
  "robustbase", "ordinal", "MASS", "data.table", "future", "future.apply", 
  "SimDesign",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

>>>>

# Automatically create a bib database for R packages
packagesForBibliography <- c(
  "petersenlab", "ffanalytics",
  "knitr", "rmarkdown", "broom", "broom.mixed", "psych", "downlit", "xml2", 
  "gsisdecoder", "progressr", "DescTools", "pwr", "pwrss", "WebPower", "XICOR", 
  "dagitty", "ggdag", "ggtext", "gghighlight", "ggExtra", "grid", "patchwork", 
  "pROC",  "lme4", "lmerTest", "MuMIn", "emmeans", "pbkrtest", "sjstats", 
  "AICcmodavg", "rstan", "brms", "tidybayes", "bbmle", "fitdistrplus", "sn", 
  "mclust", "magrittr", "viridis", "viridisLite", "msir", "plotly", "webshot2", 
  "quantmod", "fPortfolio", "NMOF", "nFactors", "xts", "zoo", "forecast",
  "stringi", "parallelly", "doParallel", "missRanger", "ggridges", "powerjoin", 
  "caret", "LongituRF", "gpboost", "cmdstanr", "corrplot", "mgcv", "rms", 
  "car", "lavaan", "lavaanPlot", "mice", "miceadds", "interactions", 
  "robustbase", "ordinal", "MASS", "data.table", "future", "future.apply", 
  "SimDesign", "domir",
  "tidyverse", "ggplot2", "dplyr", "lubridate", # packages that are part of the tidyverse ecosystem
  "nflverse", "nflfastR", "nflseedR", "nfl4th", "nflreadr", "nflplotR", # packages that are part of the nflverse ecosystem
  "tidymodels", "rsample", "parsnip", "recipes", "tune", "yardstick", # packages that are part of the tidymodels ecosysyem
  "easystats", "see", "performance", "parameters", "correlation", "effectsize", # packages that are part of the easystats ecosystem
  "learnr", "remotes", "shiny", "PearsonDS", "lavaangui", "dynamic" # packages that are cited but not used for computation (and thus do not need to be installed by me or readers)
)

### 2025-07-12 multiple-regression.qmd
20250712 - add interpretation

The *standard error* of a regression coefficient represents the uncertainty of the parameter.
If we have less uncertainty (i.e., more confidence) about the parameter, the standard error will be small.
If we have more uncertainty (i.e., less confidence) about the parameter, the standard error will be large.
If we used the same sampling procedure repeatedly and calculated the regression coefficient each time, the true parameter in the population would fall 68% of the time within the interval of:  $[\text{model parameter estimate for the regression coefficient} \pm 1 \text{ standard error}]$ .

>>>>

The *standard error* of a regression coefficient represents the imprecision or uncertainty of the parameter.
If we have less uncertainty (i.e., more confidence) about the parameter, the standard error will be small, reflecting greater precision of the regression coefficient.
If we have more uncertainty (i.e., less confidence) about the parameter, the standard error will be large, reflecting less precision of the regression coefficient.
If we used the same sampling procedure repeatedly and calculated the regression coefficient each time, the true parameter in the population would fall 68% of the time within the interval of:  $[\text{model parameter estimate for the regression coefficient} \pm 1 \text{ standard error}]$.
The standard error is related to the sample size—the larger the sample size, the smaller the standard error (the greater the precision of our estimate of the regression coefficient).
Otherwise said, having more data gives more precise estimates and thus increases [statistical power](#sec-statisticalPower).

### 2025-07-12 pca.qmd
20250712 - add interpretation

## Ensuring the Variables are on the Same Scale {#sec-pcaStandardize}

Before performing a PCA, it is important to ensure that the variables included in the PCA are on the same scale.
PCA seeks to identify components that explain variance in the data, so if the variables are not on the same scale, some variables may contribute considerably more variance than others.
A common way of ensuring that variables are on the same scale is to standardize them using, for example,  [*z*-scores](#sec-zScores).

## Component Rotation {#sec-pcaRotation}

@@

## Determining the Number of Components to Retain {#sec-pcaNumComponents}

@@

## Interpreting and Using PCA Components {#sec-interpretingAndUsingComponentsPCA}

@@

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-pcaVSSPlot1] and [-@fig-pcaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].

@@

@@

>>>>

## Decisions in Principal Component Analysis {#sec-pcaDecisions}

There are four primary decisions to make in PCA:

1. what variables to include in the model and how to scale them
1. whether and how to rotate components
1. how many components to retain
1. how to interpret and use the components

As in [factor analysis](#sec-factorAnalysis), the answer you get can differ highly depending on the decisions you make.
We provide guidance on each of these decisions below and in @sec-factorAnalysisDecisions.

### 1. Variables to Include and their Scaling {#sec-variablesToInclude-pca}

As in [factor analysis](#sec-factorAnalysis).
The first decision when conducting a factor analysis is which variables to include and the scaling of those variables.
Guidance on which variables to include is in @sec-variablesToInclude-factorAnalysis.

In addition, before performing a PCA, it is important to ensure that the variables included in the PCA are on the same scale.
PCA seeks to identify components that explain variance in the data, so if the variables are not on the same scale, some variables may contribute considerably more variance than others.
A common way of ensuring that variables are on the same scale is to standardize them using, for example,  [*z*-scores](#sec-zScores).

### 2. Component Rotation {#sec-pcaRotation}

@@

### 3. Determining the Number of Components to Retain {#sec-pcaNumComponents}

@@

### 4. Interpreting and Using PCA Components {#sec-interpretingAndUsingComponentsPCA}

@@

The number of components to keep would depend on which criteria one uses.
Based on the rule to keep factors whose eigenvalues are greater than one and based on the parallel test, we would keep nine components.
However, based on the Cattell scree test (the "elbow" of the screen plot minus one), we would keep three components.
If using the optimal coordinates, we would keep four components; if using the acceleration factor, we would keep one component.
Therefore, interpretability of the components would be important for deciding how many components to keep.

@@

We generated the very simple structure (VSS) plots in Figures [-@fig-pcaVSSPlot1] and [-@fig-pcaVSSPlot2] using the `vss()` and `nfactors()` functions of the `psych` package [@R-psych].
The optimal number of components based on VSS criterion is three components.

@@

Based on the component solutions, the three-component solution maps onto our three-factor solution from [factor analysis](#sec-factorAnalysis).
The three-component solution explains more than half of the variance in the variables.
The fourth component in a four-component solution is not particularly interpretable, explains relatively little variance, and seems to be related to sacks: sacks suffered, sack yards lost, sack fumbles, and sack fumbles lost.
However, two of the sack-related variables (sacks suffered and sack yards lost) load more strongly onto the first component, suggesting that these sack-related variables are better captured by another component.
Moreover, sacks might be considered a scoring methodological component rather than a particular concept of interest.
For these resources we prefer the three-component solution and choose to retain three components.

@@

Below are component scores from the PCA for the first six players:
